\documentclass[12pt]{article}
\usepackage{multicol, parskip, times}

\title{Ziva Kwawakabva Online Shona Tutoring Notes}
\author{Mufaro Machaya}
\date{June 2025}

\newcommand{\dw}[2]{\textbf{#1}\hspace{1em}\textit{#2}\\}
\newcommand{\tl}[2]{``#1''\\\textit{#2}}
\newcommand{\resp}[4]{%
  \tl{#1}{#2} \\
  \tl{#3}{#4}
}

\begin{document}
\maketitle
\newpage
\section{Vocabulary}
\begin{multicols}{2}
\subsection{Nouns}
\subsubsection{Food}
\dw{mvura}{water}
\dw{chibage}{corn}
\dw{mbambaira}{sweet potato}
\dw{mupunga}{rice}
\dw{nzungu}{peanuts}
\dw{munyu}{salt}

\subsubsection{Relatives}
\dw{amai}{mother}
\dw{baba}{father}
\dw{sekuru}{grandfather}
\dw{ambuya}{grandmother}
\dw{mwana}{child/baby}
\dw{tete}{father's sister}
\dw{mainini}{mother's younger sister}
\dw{maiguru}{mother's older sister}
\dw{babamunini}{father's younger brother}
\dw{babamukuru}{father's older brother}
\dw{munin'ina}{younger brother or sister}
\dw{mukoma}{older brother or sister}
\dw{mukomana}{boy}
\dw{musikana}{girl}
\dw{murume}{man}
\dw{mukadzi}{woman}

\subsubsection{Animals/Creatures}
\dw{imbwa}{dog}
\dw{mombe}{cow}
\dw{hove}{fish}
\dw{mbudzi}{goat}
\dw{huku}{chicken}
\dw{mhuka}{animals}
\dw{mbongoro}{donkey}

\subsubsection{Clothes/Body Parts}
\dw{shangu/bhutsu}{shoe/shoes}
\dw{bhachi}{coat}
\dw{bhurukwa}{pants}
\dw{bhande}{belt}
\dw{mhete}{earrings}
\dw{nguwani}{hat}
\dw{tsoka}{feet}
\dw{gumbo}{leg}
\dw{maoko}{hands}
\dw{zvipfeko}{clothes}
\dw{ziso}{eye}

\subsubsection{Things you own}
\dw{mugoti}{cooking stick}
\dw{mutsvairo}{broom}
\dw{ndiro}{plate}
\dw{bhuku}{book}
\dw{bhasikoro}{bike}

\subsubsection{Occupations/People}
\dw{shamwari}{friend}
\dw{mudzidzisi}{teacher}
\dw{mukoti}{nurse}
\dw{chiremba}{doctor}
\dw{mutyairi}{driver}
\dw{moto}{fire}
\dw{mota}{car}
\dw{murimi}{farmer}
\dw{muvhimi}{hunter}
\dw{mupurisa}{policeman/woman}
\dw{gweta}{lawyer}
\dw{muimbi}{singer}
\dw{mambo}{king}

\subsubsection{Weather}
\dw{kunaya}{raining}
\dw{kupisa}{hot}
\dw{hutonhora}{cold}
\dw{zuva}{sun/day}
\dw{gore}{year/cloud}

\subsubsection{Time (Today)}
\dw{mangwanani}{morning}
\dw{masikati}{afternoon}
\dw{maneru}{evening}
\dw{nezuro}{yesterday}
\dw{nhasi}{today}

\subsubsection{Days of the Week}
\dw{muvhuro}{monday}
\dw{chipiri}{tuesday}
\dw{chitatu}{wednesday}
\dw{china}{thursday}
\dw{chishanu}{friday}
\dw{mugovera}{saturday}
\dw{svondo}{sunday}

\subsubsection{Time (Weeks)}
\dw{vhiki/svondo}{week}
\dw{vhiki rino}{this week}
\dw{vhiki rakapera}{last week}
\dw{vhiki rinouya}{next week}

\subsubsection{Time (Months)}
\dw{mwedzi}{month}
\dw{mwedzi uno}{this month}
\dw{mwedzi wakapera}{last month}
\dw{mwedzi unouya}{next month}

\subsubsection{Time (Years)}
\dw{gore}{year}
\dw{gore rino}{this year}
\dw{gore rakapera}{last year}
\dw{gore rinouya}{next year}

\subsubsection{Other}
\dw{zororo}{break}
\dw{basa}{work}
\dw{chikoro}{school}
\dw{chitoro}{store}
\dw{maruva}{flowers}


\subsection{Verbs}
\dw{bika}{cook}
\dw{bata}{touch/hold/catch}
\dw{rara}{sleep}
\dw{dira}{pour}
\dw{muka}{wake up}
\dw{fara}{be happy}
\dw{ona}{see}
\dw{buda}{go out}
\dw{gara}{sit}
\dw{ifa}{die}
\dw{ipa}{give}
\dw{ida}{like}
\dw{isa}{put}
\dw{seka}{laugh}
\dw{geza}{bath/wash}
\dw{vaka}{build}
\dw{tsvaga/wana}{find}
\dw{simuka}{stand up}
\dw{tarisa}{look}
\dw{tevera}{follow}
\dw{buruka}{get down}
\dw{kurumidza}{hurry/be quick}
\dw{dzinga}{chase away}
\dw{nhonga}{pick up}
\dw{shaya}{lack}
\dw{unza}{bring}
\dw{shevedza}{call}
\dw{dimbura}{break}
\dw{rasa}{throw away}
\dw{ziva}{know}
\dw{namata}{pray}
\dw{uya}{come}
\dw{mira}{stand/wait}
\dw{famba}{walk}
\dw{mhanya}{run}
\dw{dzidza}{learn}
\dw{verenga}{count/read}
\dw{nyora}{write}
\dw{pinda}{come in}
\dw{simudza}{lift up}
\dw{idya}{eat}
\dw{inwa}{drink}
\dw{pedza}{finish}
\dw{chema}{cry}
\dw{dhidha}{swim}
\dw{nyarara}{be quiet}
\dw{bhururuka}{fly}
\dw{tenga}{buy}
\dw{tengesa}{sell}
\dw{tanga}{start/begin}
\dw{vhara}{close}
\dw{vhura}{open}
\dw{tamba}{play/dance}
\dw{enda}{go}
\dw{pfeka}{wear/put on}
\dw{cheka}{cut}
\dw{pfura}{shoot}
\dw{udza}{tell/inform}
\dw{batsira}{help}
\dw{gadzira}{fix/repair}
\dw{nongedza}{point to}

\subsection{Negatives}
\dw{Handidi}{I don't want}
\dw{Handina}{I don't have}
\dw{Handizivi}{I don't know}
\dw{Handikudi}{I don't love you}
\dw{Handiende}{I don't go}
\dw{Handisi}{I am not}
\dw{Hausi}{You are not}
\dw{Hatisi}{We are not}
\dw{Hamusi}{You are not (plural)}
\dw{Handidyi}{I don't eat}
\dw{Handinwi}{I don't drink}
\dw{Handigezi}{I don't bathe}
\dw{Haarare}{He/she does not sleep}

\subsection{Phrases/Sentence Fragments}

\subsubsection{Common Phrases}
\dw{nguva dzose}{all the time}
\dw{mazuva ose}{everyday}
\dw{mazuvano}{these days}
\dw{roga-roga}{each and every}
\dw{mangwanani ano}{this morning}
\dw{kuma}{around}
\dw{manje-manje}{soon}
\dw{marimwezuro}{two days ago}

\subsubsection{Numbers}
\dw{motsi}{one}
\dw{piri}{two}
\dw{tatu}{three}
\dw{ina}{four}
\dw{shanu}{five}
\dw{tanhatu}{six}
\dw{nomwe}{seven}
\dw{sere}{eight}
\dw{pfumbamwe}{nine}
\dw{gumi}{ten}

\end{multicols}

\section{Full Conversation}

\subsection{Greetings}
\subsubsection{First/Long Time Greeting (formal)}
\tl{Makadii henyu <A>.}{}
\tl{Ndiripo, makadiiwo?}{}
\tl{Ndidipo.}{}

\subsubsection{First/Long Time Greeting (informal)}
\tl{Wakadii hako <A>?}{}
\tl{Ndiripo wakadiiwo?}{}
\tl{Ndiripo.}{}

\subsubsection{Morning (formal)}
\tl{Mangwanani <A>.}{}
\tl{Mangwanani <B>, mamuka sei?}{}
\tl{Ndamuka mamukawo}{}
\tl{Ndamuka}{}

\subsubsection{Morning (informal)}
\tl{Mangwanani <A>.}{}
\tl{Mangwanani <B>, wamuka sei?}{}
\tl{Ndamuka.}{}

\subsubsection{Afternoon (formal)}
\tl{Masikati <A>}{}
\tl{Masikati <B>, maswera sei?}{}
\tl{Ndaswera maswerawo.}{}
\tl{Ndaswera.}{}

\subsubsection{Afternoon (informal)}
\tl{Masikati <A>.}{}
\tl{Masikati <B>, waswera sei?}{}
\tl{Ndaswera.}{}

\subsubsection{Evening (formal)}
\tl{Manheru <A>}{}
\tl{Manheru <B>, maswera sei?}{}
\tl{Ndaswera maswerawo.}{}
\tl{Ndaswera.}{}

\subsubsection{Evening (informal)}
\tl{Manheru <A>}{}
\tl{Manheru <B>, waswera sei?}{}
\tl{Ndawera.}{}

\subsection{About You}
\begin{enumerate}
\item \resp{Unonzi ani?}{What is your name?}{Ndinonzi Mufaro.}{My name is Mufaro.}
\item \resp{Unogara kupi?}{Where do you live?}{Ndinogara kuHouston}{I live in Houston.}
\item \resp{Unogara nani?}{Who do you live with?}{Ndinogara na amai na baba na Tinofara}{I live with my mother, my father, and Tinofara.}
\item \resp{Une makore mangani}{How old are you?}{Ndine makore gumi nemapfumbamwe gore.}{I am nineteen years old.}
\item \resp{Chii chaunofarira kudya?}{What do you like to eat?}{Ndinofarira kudya nyama yemombe ne mupanga uye inwa mvura kutonhora.}{I like to eat beef with rice and drink cold water.}
\item \resp{Chii chaunofarira kuita?}{What do you like to do}{Ndinofarira kugadzira mota yangu uye kurara zvakanaka.}{I like to fix my car and sleep}
\item \resp{Chikoro chako chinonzi chii?}{What is the name of your school?}{Chikoro chanugu chinonzi UBC.}{The name of my school is UBC.}
\item \resp{Mudzidzisi wako anonzi ani?}{What is the name of your teacher?}{Mudzidzisi wangu anonzi John.}{My teacher is named John.}
\item \resp{Shamwari yako inonzi ani?}{What is your friend's name?}{Shamwari yangu inonzi Kevin.}{My friend is named Kevin.}
\item \resp{Hanzvadzi yako ndiani?}{Who is your brother if you are female or sister if you are male?}{Handina hanzvadzi.}{I do not have a sister.}
\item \resp{Munin'ina wako anonzi ani?}{What is the name of your younger sister (if you are female) or younger brother (if you are male)?}{Handina munin'ina.}{I don't have a younger brother.}
\item \resp{Mukomana wako ndiani?}{Who is your older brother (if you are male) or older sister (if you are female)?}{Mukomana wangu ndiTinofara.}{My older brother is named Tinofara.}
\item \resp{Kumusha kwenyu ndekupi?}{Where is your rural home?}{Kumusha kwedu kuri kunze kweGweru.}{Our home is outside of Gweru.}
\item \resp{Ambuya vako vanonzi ani?}{What is your grandmother's name?}{Ambuya vangu vanonzi Jane.}{My grandmother's name is Jane.}
\item \resp{Sekuru vako vanonzi ani?}{What is your grandfather's name?}{Handizivi anonzi Sekuru vangu.}{I don't know what my grandfather's name is.}
\item \resp{Unoerei?}{What is your totem?}{Ndinoera shumba.}{My totem is the lion.}
\item \resp{Musikana/Mukomana wako ndiani?}{Who is your girlfriend/boyfriend?}{Handina musikana.}{I don't have a girlfriend.}
\item \resp{Unoshanda kupi?}{Where do you work?}{Handiende basa.}{I don't go to work.}
\end{enumerate}

\section{June 4}
\subsection{Pre-Class Homework}
\begin{enumerate}
\item \tl{Handidi kupa Tinofara bhachi rangu!}{I don't want to give Tinofara my coat!}
\item \tl{Handina nguva kurumidza kutaura.}{I don't have time to talk, hurry up.}
\item \tl{Handizivi kugadzira mota nokuti handina kuwuna nguva yekudzidzira kugadzira motakari.}{I don't know how to fix my car because I don't have the time to learn repair.}
\item \tl{Handikudi nokuti hanisi kufara.}{I don't love you because I am not happy.}
\item \tl{Handiende kuDeutschland nokuti handi taurai Deutsch!}{I don't go to Germany because I don't speak German.}
\item \tl{Nasi, handisi kuenda kuchikoro nokuti kuri kutonhora ne kunaya.}{Today, I will not go to school because it is raining and cold.}
\item \tl{Handidyi mombe ne mbudzi ne huku ne hove ne mbongoro ne imbwa nokuti ndiri vegetarian.}{I don't eat cow, goat, chicken, fish, donkey, and dog because I'm vegetarian.}
\item \tl{Handinwi mvura inopisa.}{I don't drink hot water.}
\item \tl{Handigezi kuchikoro.}{I don't bathe at school.}
\item Haarare means
\end{enumerate}

\subsection{Post-Class Homework}
\begin{enumerate}
\item \tl{Handidi kuenda kuchikoro.}{I don't want to go to school.}
\item \tl{Handina marii ye zvipfeko.}{I don't have enough money for the clothes.}
  \begin{enumerate}
  \item \tl{Anoita marii madomasi?}{How much are the tomatoes?}
  \end{enumerate}
\item \tl{Handizivi anonzi yako\footnote{Missing: who's name it is?}.}{I don't know the name.}
\item \tl{Handikudi nhasi.}{I don't love you today.}
\item \tl{Handiende kubasa mangwanani ano.}{I don't go to work this morning.}
\item \tl{Handisi mutambi.}{I am not an actor/playing.}
\item \tl{Handidyi nguruve.}{I don't eat pork.}
\item \tl{Handinwi Doro asi ndanwa mukaka.}{I don't drink beer but I drink milk.}
\item \tl{Handigezi manheru.}{I don't bathe in the evening.}
\end{enumerate}

\section{June 11}
\subsection{Tenses}
Tenses in shona are enoted by a specific sound preceding a verb (either as a standalone word or as a prefix):
\begin{itemize}
\item ``ka'' for the distant past (yesterday and earlier).
\item There is no special prefix for things that happened today, but not the immediate present. (default).
\item ``iri'' for the present.
\item ``cha'' for the future.
\end{itemize}

\subsubsection{Distant Past Tense Examples (ka)}
\begin{enumerate}
\item \dw{kudya}{eat} \tl{Nezuro, nda\textbf{ka}dya sadya.}{Yesterday I ate sadza.}
\item \dw{kugeza}{bathe} \tl{Gore rakapera, ta\textbf{ka}geza kuhotera.}{Last year, we bathed at the hotel.}
\item \dw{kumhanya}{run} \tl{Baka va\textbf{ka}mhanya kuchitoro.}{Dad ran to the store.}
\item \dw{kutaura}{talk/speak} \tl{Sekuru va\textbf{ka}taura nagogo.}{Grandpa talked to grandma.}
\item \dw{kufamba}{walk} \tl{Nda\textbf{ka}famba kuenda kuchikoro nezuro.}{I walked to school yesterday.}
\item \dw{kunwa}{drink} \tl{Gore rakapera, nda\textbf{ka}nwa Mazoe Orange akawanda, asi zvino handinwi Mazoe nokuti handina mari yakakwana yekutenga Mazoe kuchitoro.}{Last year, I drank a lot of Mazoe Orange, but now, I don't drink Mazoe because I don't have enough money to buy Mazoe at the store.}
\item \dw{kuseka}{laugh} \tl{Handina kuseka mwedzi wakapera uye uno nokuti ambuya vangu vakafa.}{I did not laugh last month and this month because my grandmother died.}
\item \dw{kuenda}{} \tl{Vhiki rakapera, ndakaenda kubasa kwandino gadzira mota dzakawanda.}{Last week, I went to work where I fix many cars.}
\end{enumerate}

\subsubsection{Past (Today) Examples (no prefix)}
\begin{enumerate}
\item \dw{kudya}{eat} \tl{Mangwanani ndadya sadza.}{In the morning, I ate sadza.}
\item \dw{kugeza}{bathe} \tl{Nhasi tageza kuhotera.}{Today we bathed at the hotel.}
\item \dw{kumhanya}{run} \tl{Baba vamhanya kuchitoro.}{Dad ran to the store.}
\item \dw{kutaura}{talk/speak} \tl{Nhasi, ndafonera shamwari yangu anonzi David tikataura nezve chikoro nebasa.}{Today, I called my friend named David and we spoke about school and work.}
\item \dw{kufamba}{walk} \tl{Masikati ndafamba kuenda kubank kunotanga kutengesa imba yangu.}{In the evening, I walked to the bank to begin selling my house.}
\item \dw{kunwa}{drink} \tl{Ndichango pedza kunwa komichi yemukaka ne kudya kwangu kwemasikati.}{I just finished drinking a cup of milk with my lunch.}
\item \dw{kuseka}{laugh} \tl{Ndaseka zvakanyanya nokuti shamwari yangu yataura joke inosekesa.}{I just laughed a lot because my friend said a funny joke.}
\end{enumerate}

\subsubsection{Present Tense Examples (iri)}
\begin{enumerate}
\item \dw{kudya}{eat} \tl{\textbf{Ndiri} kudya sadza.}{I am eating sadza.}
\item \dw{kugeza}{bathe} \tl{\textbf{Tiri} kugeza kuhotera.}{We are bathing at the hotel.}
\item \dw{kumhanya}{run} \tl{Baba \textbf{vari} kumhanya kuchitoro.}{Dad is running to the store.}
\item \dw{kutaura}{talk/speak}
\item \dw{kufamba}{walk}
\item \dw{kunwa}{drink}
\item \dw{kuseka}{laugh}
\end{enumerate}

\subsubsection{Future Tense Examples (cha)}
\begin{enumerate}
\item \dw{kudya}{eat} \tl{Mangwana, ndi\textbf{cha}dya}{Tomorrow, I will eat sadza.}
\item \dw{kugeza}{bathe} \tl{Manheru, ti\textbf{cha}geza kuhotera.}{In the evening, we will bathe in the hotel.}
\item \dw{kumhanya}{run} \tl{Baba va\textbf{cha}mhanya kuchitoro.}{Dad will run at the store.}
\item \dw{kutaura}{talk/speak}
\item \dw{kufamba}{walk}
\item \dw{kunwa}{drink}
\item \dw{kuseka}{laugh}
\end{enumerate}

\subsection{Conjunctions}

\end{document}
